The Power of Unbiased Recursive Partitioning:\\ A Unifying View of CTree, MOB, and GUIDE

### 2019-03-29 pkg/partykit/inst/guideliketest/draft/draft.Rnw
added evaluation with continuous coefficients to appendix; changed color palletes; further minor revisions, also based on feedback from DAGStat2019 presentation
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2506 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: b1baf05c6a93261edf18f4ce93de579b7dde02cd

library("partykit")
library("Formula")
library("parallel")
library("lattice")
library("scales")

@@

## HCL palette
pal <- hcl(c(140, 80, 260, 70, 10), 300, 60)
names(pal) <- c("CTree", "CTree_max", "MOB", "GUIDE", "GUIDE_scores")

@@

As explained in Section~\ref{sec:introduction} the considered tree algorithms CTree, MOB, and 
GUIDE first select the split variable and then, in a separate step, the split point. 
For the first step of selecting a split variable they all apply statistical tests
following the same general strategy: 
\begin{enumerate}
\item Based on the discrepancy measure an $n \times k$-matrix $D$ is obtained giving information
on the goodness of fit of the model for a data set of $n \in \mathbb{N}$ observations in the 
current node and the parameter $k \in \mathbb{N}$ defined by the algorithm.
\item Using this matrix the algorithm-specific test statistic is evaluated for each possible 
split variable $Z_j$, $j \in \{1,\ldots,m\}$ for $m \in \mathbb{N}$, assessing the dependency
of $D$ on $Z_j$.
\item For the resulting values of the test statistics the corresponding $p$-values are calculated
in order to allow for their comparison on a standardized/unified scale and in that way for
an unbiased variable selection.
\item The variable corresponding to the smallest $p$-value is selected as split variable as 
it has shown to be the variable with the strongest influence on the goodness of fit of the model.
\end{enumerate}

@@

Even though CTree, MOB, and GUIDE differ in the specific tests they apply (see second column of
Table~\ref{tab:combinations}) their approaches on how to select a split variable follow the same
basic structure as explained in Section~\ref{sec:generic_alg}. 
Thus they can be presented together in a unifying framework for testing strategies
in model-based unbiased recursive partitioning. 
In this framework each specific testing strategy can be obtained by combining the provided 
building blocks accordingly. But apart from that, variations of the commonly known testing 
strategies can be obtained as this general toolbox allows for all possible combinations of the 
building blocks. Moreover, by exchanging only one of them its impact can be investigated in detail.
These building blocks are explained in Section~\ref{sec:buildingblocks} and their setting
for CTree, MOB, and GUIDE are listed in Table~\ref{tab:buildingblocks} which gives a more 
detailed overview of the differences in the second column of Table~\ref{tab:combinations}.

@@

\textit{Note:}
Apart from these three building blocks, the algorithms differ in their approximation of the 
underlying null distribution of the test statistic. While the permutation tests in the CTree 
algorithm work in a conditional framework the MOB and GUIDE algorithms both apply an unconditional 
approach.

\begin{table}[t]
\centering
\caption[Table caption text]{Testing strategies of CTree, MOB, and GUIDE with the corresponding
setting of the building blocks in the unifying framework and the approximation type of the
underlying null distribution.}
\label{tab:buildingblocks}
\begin{tabular}{ l c c c c }
\hline 
      & Scores            & Binarization  & Categorization  & Null distribution \\ 
\hline
CTree & full model scores & --          & --            & conditional\\
MOB   & full model scores & --          & --            & unconditional\\
GUIDE & residuals         & \checkmark  & \checkmark    & unconditional\\
\hline
\end{tabular}
\end{table}

@@

\subsection{Linear Model Tree}\label{sec:lmtrees}
As an example a linear model tree is considered as this allows for an easy and well structured 
description of the investigated subjects. However, the conclusions drawn from this example also 
hold for a wide range of other model types. 
The models fitted in the nodes of the linear model tree as considered for the simulation 
study in Section~\ref{sec:simulation} are of the form
$$
Y = \beta_0 + \beta_1 \cdot X + \epsilon
$$
for a response variable $Y$, a regressor variable $X$ and an error $\epsilon$. 
For each node the type of the fitted model is the same such that no node-specific features are 
considered and the observations of the current (arbitrarily chosen) node are denoted by 
$\{(y_i,x_i)\}_{i=1,\ldots,n}$ for $n \in \mathbb{N}$.

A very basic form of a discrepancy measure can be obtained for each observation by simply 
calculating deviations of the response values from the mean value within the node, for example 
by $(\bar{y} - y_i)$.
However, this includes only the response variable $Y$ but not the fitted linear model.
For a model-based approach residuals can be considered defined by
$$
r(y_i,x_i,\beta_0,\beta_1) = y_i - \beta_0 - \beta_1 \cdot x_i.
$$
The appropriate choice of the function $\psi$ is then
$$
\psi(y_i,x_i,\beta_0,\beta_1) = r(y_i,x_i,\beta_0,\beta_1)^2
$$
such that $\sum_{i=1}^n \psi(y_i,x_i,\beta_0,\beta_1)$ is the objective function
which is minimized to obtain the estimated coefficient vector
$\boldsymbol{\hat{\beta}} = (\hat{\beta_0}, \hat{\beta_1})^\top$.
The corresponding score function is defined by
$$
s(y_i, x_i, \beta_0, \beta_1) = 
\frac{\partial \psi}{\partial \boldsymbol{\beta}}(y_i, x_i, \beta_0, \beta_1) = 
2 \cdot r(y_i,x_i, \beta_0, \beta_1) \cdot (1,x_i)^\top
$$
where the factor $2$ is often omitted such that the first column of the $n \times 2$-dimensional
score matrix corresponds to the residuals. 
Therefore, including only residuals (i.e. only the first column of the score matrix)
corresponds to including the derivatives of $\psi$ only regarding the intercept 
parameter~$\beta_0$.
On the contrary, by using full model scores (i.e. the whole $n \times 2$ matrix) 
the derivatives of $\psi$ regarding the intercept $\beta_0$ (first column) and the slope 
parameter $\beta_1$ (second column) are considered.

@@

paltrees <- rgb(c(1, 0.9, 1), c(1, 0.9, 1), c(1, 0.9, 1))

ep <- function (obj, digits = 3, abbreviate = FALSE, justmin = Inf, 
                just = c("alternate", "increasing", "decreasing", "equal"), 
                fill = "white") 
{
  meta <- obj$data
  justfun <- function(i, split) {
    myjust <- if (mean(nchar(split)) > justmin) {
      match.arg(just, c("alternate", "increasing", "decreasing", "equal"))
    } else {
      "equal"
    }
    k <- length(split)
    rval <- switch(myjust, equal = rep.int(0, k), 
                   alternate = rep(c(0.5, -0.5), length.out = k), 
                   increasing = seq(from = -k/2, to = k/2, by = 1), 
                   decreasing = seq(from = k/2, to = -k/2, by = -1))
    unit(0.5, "npc") + unit(rval[i], "lines")
  }
  function(node, i) {
    split <-  c("<= xi", "> xi")
    y <- justfun(i, split)
    split <- split[i]
    if (any(grep(">", split) > 0) | any(grep("<", split) > 0)) {
      tr <- suppressWarnings(try(parse(text = paste("phantom(0)", split)), silent = TRUE))
      if (!inherits(tr, "try-error")) split <- tr
    }
    grid.rect(y = y, gp = gpar(fill = fill, col = 0), width = unit(1, "strwidth", split))
    grid.text(split, y = y, just = "center")
  }
}
class(ep) <- "grapcon_generator"
@

\begin{figure}%[t]
\setkeys{Gin}{width=1\linewidth}
\minipage{0.45\textwidth}
<<dgp_stump1, fig=TRUE, echo=FALSE, width=4, height=3>>=
#par(mar=c(3,3,0,0), oma = c(3,3,0,0))
plot(p, 
     tp_args = list(FUN = identity, width = 15, fill = paltrees[c(1, 3)]), 
     ip_args = list(fill = paltrees[c(2, 3)]),
     drop_terminal = TRUE, tnex = 1, edge_panel = ep)
@
\endminipage
\minipage{0.45\textwidth}
<<dgp_stump2, fig=TRUE, echo=FALSE, width=4, height=3>>=
par(mar = c(4,4,2,4))
plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x", 
     main = expression("varying"~beta[0]),
     pch = 19, col = alpha("blue", 0.4), cex = 0.7)
abline(coef = c(-1,1), col = "blue", lwd = 2)
abline(coef = c(1,1), col = "red", lwd = 2)
points(x = d$x[d$z1>xi], y = d$y[d$z1>xi], ylim = c(-5,5), pch = 19, col = alpha("red", 0.4), cex = 0.7)
legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c("blue", "red"), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==+delta), side = 4, col = "red", las = 1, line = 0.5, at = 3.2)
mtext(expression(beta[1]== 1), side = 4, col = "red", las = 1, line = 0.5, at = 2.2)
mtext(expression(beta[0]==-delta), side = 4, col = "blue", las = 1, line = 0.5, at = 0.2)
mtext(expression(beta[1]== 1), side = 4, col = "blue", las = 1, line = 0.5, at = -0.8)
@
\endminipage

@@

par(mar = c(4,4,2,4))
plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x", main = expression("varying"~beta[1]),
     pch = 19, col = alpha("blue", 0.4), cex = 0.7)
abline(coef = c(0,1), col = "blue", lwd = 2)
abline(coef = c(0,-1), col = "red", lwd = 2)
points(x = d$x[d$z1>xi], y = d$y[d$z1>xi], ylim = c(-5,5), pch = 19, col = alpha("red", 0.4), cex = 0.7)
legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c("blue", "red"), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==0), side = 4, col = "red", las = 1, line = 0.5, at = -1)
mtext(expression(beta[1]==-delta), side = 4, col = "red", las = 1, line = 0.5, at = -2)
mtext(expression(beta[0]==0), side = 4, col = "blue", las = 1, line = 0.5, at = 2)
mtext(expression(beta[1]==+delta), side = 4, col = "blue", las = 1, line = 0.5, at = 1)
@
\endminipage
\minipage{0.45\textwidth}
<<dgp_stump4, fig=TRUE, echo=FALSE, width=4, height=3>>=
d <- dgp_stump(nobs = nobs, binary_beta = TRUE, binary_regressor = FALSE, only_intercept = FALSE,
               vary_beta = "all", delta = delta, xi = xi, sigma = sigma)
par(mar = c(4,4,2,4))
plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x",
     main = expression("varying"~beta[0]~"and"~beta[1]),
     xlim = c(-1,1.1),
     pch = 19, col = alpha("blue", 0.4), cex = 0.7)
abline(coef = c(-1,1), col = "blue", lwd = 2)
abline(coef = c(1,-1), col = "red", lwd = 2)
points(x = d$x[d$z1>xi], y = d$y[d$z1>=xi], pch = 19, col = alpha("red", 0.4), cex = 0.7)
legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c("blue", "red"), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==-delta), side = 4, col = "blue", las = 1, line = 0.5, at = 1.7)
mtext(expression(beta[1]==+delta), side = 4, col = "blue", las = 1, line = 0.5, at = 0.7)
mtext(expression(beta[0]==+delta), side = 4, col = "red", las = 1, line = 0.5, at = -0.7)
mtext(expression(beta[1]==-delta), side = 4, col = "red", las = 1, line = 0.5, at = -1.7)
@
\endminipage
\caption{\label{fig:dgp_stump}{Top left panel: True stump structure with the true model parameters $\beta_0$ and $\beta_1$, either varying or being fixed at $0$ or $1$ respectively.
Top right and bottom panels: Bivariate plot of the response $Y$ on regressor $X$ 
with effect size $\delta = 1$ and for three cases: 
varying $\beta_0$ and fixed $\beta_1$ (top right),
fixed $\beta_0$ and varying $\beta_1$ (bottom left),
varying $\beta_0$ and varying $\beta_1$ (bottom right).}}
\end{figure}

@@

\begin{figure}%[t]
\begin{center}
\setkeys{Gin}{width=1\linewidth}
\minipage{0.5\textwidth}
<<dgp_tree1, fig=TRUE, echo=FALSE, width=5, height=4.5>>=
par(mar=c(0,0,0,0), oma=c(0,0,0.4,0))
#paltrees <- rgb(c(0.97, 0.64, 1), c(0.70, 0.83, 1), c(0.30, 0.99, 1))
paltrees <- rgb(c(1, 0.9, 1), c(1, 0.9, 1), c(1, 0.9, 1))
plot(p, edge_panel = ep,
     tp_args = list(FUN = identity, width = 15, fill = paltrees[c(1, 3)]), 
     ip_args = list(fill = paltrees[c(2, 3)]),
     drop_terminal = TRUE, tnex = 1)
@
\endminipage
\minipage{0.45\textwidth}
<<dgp_tree2, fig=TRUE, echo=FALSE, width=4, height=4>>=
par(mar = c(4,4,2,4))
plot(x = d$x[d$z2<=xi], y = d$y[d$z2<=xi], ylim = c(-5,5), ylab = "y", xlab = "x",
     pch = 19, col = alpha("blue", 0.4), cex = 0.7)
abline(coef = c(0,1), col = "blue", lwd = 2)
abline(coef = c(-1,-1), col = "red", lwd = 2)
abline(coef = c(1,-1), col = "darkgreen", lwd = 2)
points(x = d$x[d$z2>xi & d$z1<=xi], y = d$y[d$z2>xi & d$z1<=xi], pch = 19, 
       col = alpha("red", 0.4), cex = 0.7)
points(x = d$x[d$z2>xi & d$z1>xi], y = d$y[d$z2>xi & d$z1>xi], pch = 19, 
       col = alpha("darkgreen", 0.4), cex = 0.7)
legend("top", c(expression(z[2]<=xi), 
                expression(z[2]>xi~" and "~z[1]<=xi),
                expression(z[2]>xi~" and "~z[1]>xi)), 
       col = c("blue", "red", "darkgreen"), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==0), side = 4, col = "blue", las = 1, line = 0.5, at = 2)
mtext(expression(beta[1]==+delta), side = 4, col = "blue", las = 1, line = 0.5, at = 1.3)
mtext(expression(beta[0]==+delta), side = 4, col = "darkgreen", las = 1, line = 0.5, at = 0.1)
mtext(expression(beta[1]==-delta), side = 4, col = "darkgreen", las = 1, line = 0.5, at = -0.6)
mtext(expression(beta[0]==-delta), side = 4, col = "red", las = 1, line = 0.5, at = -2)
mtext(expression(beta[1]==-delta), side = 4, col = "red", las = 1, line = 0.5, at = -2.7)
@
\endminipage
\caption{\label{fig:dgp_tree}Left panel: True tree structure for the second scenario with the 
true model parameters $\beta_0$ and $\beta_1$ for each terminal panel.
Right panel: Bivariate plot of the response $Y$ on regressor $X$ 
with effect size $\delta = 1$.}
\end{center}
\end{figure}

@@

Looking at the results for the split point being close to the upper edge of the considered
interval ($\xi=0.8$, lower panels) all of the applied strategies perform worse than for the
central split point. In this case MOB is clearly ahead leading to the best results. CTree 
performs second best while both GUIDE versions struggle to detect the correct split variable 
even for a high effect size.

>>>>

library("partykit")
library("Formula")
library("parallel")
library("lattice")
library("scales")
library("colorspace")

@@

## HCL palette
pal <- qualitative_hcl(5, "Dark 3")
names(pal) <- c("CTree", "MOB", "GUIDE", "GUIDE_scores", "CTree_max")

## HCL palette for illustrations of data generating process
pal_dgp_light <- qualitative_hcl(3, "Set 2")
pal_dgp <- qualitative_hcl(3, "Dark 3")

## HCL palette for tree panels
paltrees <- rgb(c(0.9, 1), c(0.9, 1), c(0.9, 1))

@@

As explained in Section~\ref{sec:introduction} the considered tree algorithms CTree, MOB, and 
GUIDE first select the split variable and then, in a separate step, the split point. 
For the first step of selecting a split variable they all apply statistical tests
following the same general strategy: 
\begin{enumerate}
\item A discrepancy measure is obtained in the form of an $n \times k$-matrix $D$ giving information
on the goodness of fit of the model for a data set of $n \in \mathbb{N}$ observations in the 
current node and the parameter $k \in \mathbb{N}$ defined by the algorithm.
\item Using this matrix the algorithm-specific test statistic is evaluated for each possible 
split variable $Z_j$, $j \in \{1,\ldots,m\}$ for $m \in \mathbb{N}$, assessing the dependency
of $D$ on $Z_j$.
\item For the resulting values of the test statistics the corresponding $p$-values are calculated
in order to allow for their comparison on a standardized/unified scale and in that way for
an unbiased variable selection.
\item The variable corresponding to the smallest $p$-value is selected as split variable as 
it has shown to be the variable with the strongest influence on the goodness of fit of the model.
\end{enumerate}

@@

Even though CTree, MOB, and GUIDE differ in the specific tests they apply (see second column of
Table~\ref{tab:combinations}) their approaches on how to select a split variable follow the same
basic structure as explained in Section~\ref{sec:generic_alg}. 
Thus they can be presented together in a unifying framework for testing strategies
in model-based unbiased recursive partitioning. 
In this framework each specific testing strategy can be obtained by combining the provided 
building blocks accordingly. But apart from that, variations of the commonly known testing 
strategies can be obtained as this general toolbox allows for all possible combinations of the 
building blocks. Moreover, by exchanging only one of them its impact can be investigated in detail.
These building blocks are explained in Section~\ref{sec:buildingblocks} and their setting
for the standard setup of CTree, MOB, and GUIDE are listed in Table~\ref{tab:buildingblocks} which 
gives a more detailed overview of the differences in the second column of Table~\ref{tab:combinations}.

@@

\textit{Note:}
Apart from these three building blocks, the algorithms CTree, MOB, and GUIDE differ in the type of test 
statistic they apply and their approximation of the underlying null distribution of the corresponding 
test statistic. While the permutation tests in CTree work in a conditional framework both MOB (M-fluctuation tests) 
and GUIDE ($\chi^2$-tests) apply an unconditional approach. Furthermore, the algorithms CTree and MOB were 
introduced employing prepruning while GUIDE originally applies a postpruning strategy.
However, these differences refer to the standard versions of the three algorithms and all of them
allow for certain modifications of this standard setup. For example, a different form of test statistic
can be selected in CTree considering a maximum value as in MOB.

\begin{table}[t]
\centering
\caption[Table caption text]{Testing strategies of CTree, MOB, and GUIDE with the corresponding
setting of the building blocks in the unifying framework and the type of test statistic.}
\label{tab:buildingblocks}
\begin{tabular}{ l c c c c }
\hline 
      & Scores       & Binarization  & Categorization  & Statistic\\
\hline
CTree & Model scores & --	           & -- 	           & Sum of squares\\
MOB   & Model scores & --	           & -- 	           & Maximally selected\\
GUIDE & Residuals    & \checkmark    & \checkmark      & Sum of squares\\
\hline
\end{tabular}
\end{table}

@@

\subsection{Linear Model Tree}\label{sec:lmtrees}
As an example a linear model tree is considered as this allows for an easy and well structured 
description of the investigated subjects. However, the conclusions drawn from this example also 
hold for a wide range of other model types. 
The models fitted in the nodes of the linear model tree as considered for the simulation 
study in Section~\ref{sec:simulation} are of the form
$$
Y = \beta_0 + \beta_1 \cdot X + \epsilon
$$
for a response variable $Y$, a regressor variable $X$ and an error $\epsilon$. 
For each node the type of the fitted model is the same such that no node-specific features are 
considered and the observations of the (arbitrarily chosen) current node are denoted by 
$\{(y_i,x_i)\}_{i=1,\ldots,n}$ for $n \in \mathbb{N}$.

A very basic form of a discrepancy measure can be obtained for each observation by simply 
calculating deviations of the response values from the mean value within the node, for example 
by $(\bar{y} - y_i)$.
However, this includes only the response variable $Y$ but not the fitted linear model.
For a model-based approach residuals can be considered defined by
$$
r(y_i,x_i,\beta_0,\beta_1) = y_i - \beta_0 - \beta_1 \cdot x_i.
$$
The appropriate choice of the function $\psi$ is then
$$
\psi(y_i,x_i,\beta_0,\beta_1) = r(y_i,x_i,\beta_0,\beta_1)^2
$$
such that $\sum_{i=1}^n \psi(y_i,x_i,\beta_0,\beta_1)$ is the objective function
which is minimized to obtain the estimated coefficient vector
$\boldsymbol{\hat{\beta}} = (\hat{\beta_0}, \hat{\beta_1})^\top$.
The corresponding score function is defined by
$$
s(y_i, x_i, \beta_0, \beta_1) = 
\frac{\partial \psi}{\partial \boldsymbol{\beta}}(y_i, x_i, \beta_0, \beta_1) = 
-2 \cdot r(y_i,x_i, \beta_0, \beta_1) \cdot (1,x_i)^\top
$$
where the factor $-2$ is often omitted such that the first column of the $n \times 2$-dimensional
score matrix corresponds to the residuals. 
Therefore, including only residuals (i.e. only the first column of the score matrix)
corresponds to including the derivatives of $\psi$ only regarding the intercept 
parameter~$\beta_0$.
On the contrary, by using full model scores (i.e. the whole $n \times 2$ matrix) 
the derivatives of $\psi$ regarding the intercept $\beta_0$ (first column) and the slope 
parameter $\beta_1$ (second column) are considered.

@@

ep <- function (obj, digits = 3, abbreviate = FALSE, justmin = Inf, 
                just = c("alternate", "increasing", "decreasing", "equal"), 
                fill = "white") 
{
  meta <- obj$data
  justfun <- function(i, split) {
    myjust <- if (mean(nchar(split)) > justmin) {
      match.arg(just, c("alternate", "increasing", "decreasing", "equal"))
    } else {
      "equal"
    }
    k <- length(split)
    rval <- switch(myjust, equal = rep.int(0, k), 
                   alternate = rep(c(0.5, -0.5), length.out = k), 
                   increasing = seq(from = -k/2, to = k/2, by = 1), 
                   decreasing = seq(from = k/2, to = -k/2, by = -1))
    unit(0.5, "npc") + unit(rval[i], "lines")
  }
  function(node, i) {
    split <-  c("<= xi", "> xi")
    y <- justfun(i, split)
    split <- split[i]
    if (any(grep(">", split) > 0) | any(grep("<", split) > 0)) {
      tr <- suppressWarnings(try(parse(text = paste("phantom(0)", split)), silent = TRUE))
      if (!inherits(tr, "try-error")) split <- tr
    }
    grid.rect(y = y, gp = gpar(fill = fill, col = 0), width = unit(1, "strwidth", split))
    grid.text(split, y = y, just = "center")
  }
}
class(ep) <- "grapcon_generator"

tp <- function (obj, digits = 3, abbreviate = FALSE, 
                fill = c("lightgray", "white"), id = TRUE, 
                just = c("center", "top"), top = 0.85, 
                align = c("center", "left", "right"), gp = NULL, FUN = NULL, 
                height = NULL, width = NULL) 
{
  nam <- names(obj)
  extract_label <- function(node) formatinfo_node(node, FUN = FUN, 
                                                  default = c("terminal", "node"))
  maxstr <- function(node) {
    lab <- extract_label(node)
    klab <- if (is.terminal(node)) 
      ""
    else unlist(lapply(kids_node(node), maxstr))
    lab <- c(lab, klab)
    lab <- try(unlist(lapply(lab, function(x) strsplit(x, 
                                                       "\n"))), silent = TRUE)
    if (inherits(lab, "try-error")) {
      paste(rep("a", 9L), collapse = "")
    }
    else {
      return(lab[which.max(nchar(lab))])
    }
  }
  nstr <- if (is.null(width)) 
    maxstr(node_party(obj))
  else paste(rep("a", width), collapse = "")
  just <- match.arg(just[1L], c("center", "centre", "top"))
  if (just == "centre") 
    just <- "center"
  align <- match.arg(align[1L], c("center", "centre", "left", 
                                  "right"))
  if (align == "centre") 
    align <- "center"
  rval <- function(node) {
    if(node$id == 2) fill <- c("white", pal_dgp_light[1])
    if(node$id == 3 | node$id == 4) fill <- c("white", pal_dgp_light[2])
    if(node$id == 5) fill <- c("white", pal_dgp_light[3])
    fill <- rep(fill, length.out = 2)
    lab <- extract_label(node)
    if (!is.null(gp)) {
      outer_vp <- viewport(gp = gp)
      pushViewport(outer_vp)
    }
    if (is.null(height)) 
      height <- length(lab) + 1L
    node_vp <- viewport(x = unit(0.5, "npc"), y = unit(if (just == 
                                                           "top") 
      top
      else 0.5, "npc"), just = c("center", just), width = unit(1, 
                                                               "strwidth", nstr) * 1.1, height = unit(height, "lines"), 
      name = paste("node_terminal", id_node(node), sep = ""), 
      gp = if (is.null(gp)) 
        gpar()
      else gp)
    pushViewport(node_vp)
    grid.rect(gp = gpar(fill = fill[1]))
    for (i in seq_along(lab)) grid.text(x = switch(align, 
                                                   center = unit(0.5, "npc"), left = unit(1, "strwidth", 
                                                                                          "a"), right = unit(1, "npc") - unit(1, "strwidth", 
                                                                                                                              "a")), y = unit(length(lab) - i + 1, "lines"), 
                                        lab[i], just = align)
    if (id) {
      nodeIDvp <- viewport(x = unit(0.5, "npc"), y = unit(1, 
                                                          "npc"), width = max(unit(1, "lines"), unit(1.3, 
                                                                                                     "strwidth", nam[id_node(node)])), height = max(unit(1, 
                                                                                                                                                         "lines"), unit(1.3, "strheight", nam[id_node(node)])))
      pushViewport(nodeIDvp)
      grid.rect(gp = gpar(fill = fill[2], lty = "solid"))
      grid.text(nam[id_node(node)])
      popViewport()
    }
    if (is.null(gp)) 
      upViewport()
    else upViewport(2)
  }
  return(rval)
}

class(tp) <- "grapcon_generator"

\begin{figure}%[t]
\setkeys{Gin}{width=1\linewidth}
\minipage{0.45\textwidth}
<<dgp_stump1, fig=TRUE, echo=FALSE, width=4, height=3>>=
#par(mar=c(3,3,0,0), oma = c(3,3,0,0))
plot(p, 
     tp_args = list(FUN = identity, width = 15), 
     ip_args = list(fill = paltrees[c(1, 2)]),
     drop_terminal = TRUE, tnex = 1, edge_panel = ep, terminal_panel = tp)
@
\endminipage
\minipage{0.45\textwidth}
<<dgp_stump2, fig=TRUE, echo=FALSE, width=4, height=3>>=
par(mar = c(4,4,2,4))
plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x", 
     main = expression("varying"~beta[0]),
     pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
abline(coef = c(-1,1), col = pal_dgp[1], lwd = 2)
abline(coef = c(1,1), col = pal_dgp[2], lwd = 2)
points(x = d$x[d$z1>xi], y = d$y[d$z1>xi], ylim = c(-5,5), pch = 19, col = alpha(pal_dgp[2], 0.4), cex = 0.7)
legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c(pal_dgp[1], pal_dgp[2]), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==+delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = 3.2)
mtext(expression(beta[1]== 1), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = 2.2)
mtext(expression(beta[0]==-delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 0.2)
mtext(expression(beta[1]== 1), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = -0.8)
@
\endminipage

@@

par(mar = c(4,4,2,4))
plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x", main = expression("varying"~beta[1]),
     pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
abline(coef = c(0,1), col = pal_dgp[1], lwd = 2)
abline(coef = c(0,-1), col = pal_dgp[2], lwd = 2)
points(x = d$x[d$z1>xi], y = d$y[d$z1>xi], ylim = c(-5,5), pch = 19, col = alpha(pal_dgp[2], 0.4), cex = 0.7)
legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c(pal_dgp[1], pal_dgp[2]), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==0), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -1)
mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -2)
mtext(expression(beta[0]==0), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 2)
mtext(expression(beta[1]==+delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 1)
@
\endminipage
\minipage{0.45\textwidth}
<<dgp_stump4, fig=TRUE, echo=FALSE, width=4, height=3>>=
d <- dgp_stump(nobs = nobs, binary_beta = TRUE, binary_regressor = FALSE, only_intercept = FALSE,
               vary_beta = "all", delta = delta, xi = xi, sigma = sigma)
par(mar = c(4,4,2,4))
plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x",
     main = expression("varying"~beta[0]~"and"~beta[1]),
     xlim = c(-1,1.1),
     pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
abline(coef = c(-1,1), col = pal_dgp[1], lwd = 2)
abline(coef = c(1,-1), col = pal_dgp[2], lwd = 2)
points(x = d$x[d$z1>xi], y = d$y[d$z1>=xi], pch = 19, col = alpha(pal_dgp[2], 0.4), cex = 0.7)
legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c(pal_dgp[1], pal_dgp[2]), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==-delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 1.7)
mtext(expression(beta[1]==+delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 0.7)
mtext(expression(beta[0]==+delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -0.7)
mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -1.7)
@
\endminipage
\caption{\label{fig:dgp_stump}{Top left panel: True stump structure with the true model parameters $\beta_0$ and $\beta_1$, either varying or being fixed at $0$ or $1$ respectively.
Top right and bottom panels: Bivariate plot of the response $Y$ on regressor $X$ 
with effect size $\delta = 1$ and for three cases: 
varying $\beta_0$ and fixed $\beta_1$ (top right),
fixed $\beta_0$ and varying $\beta_1$ (bottom left),
varying $\beta_0$ and varying $\beta_1$ (bottom right).}}
\end{figure}

@@

\begin{figure}%[t]
\begin{center}
\setkeys{Gin}{width=1\linewidth}
\minipage{0.5\textwidth}
<<dgp_tree1, fig=TRUE, echo=FALSE, width=5, height=4.5>>=
par(mar=c(0,0,0,0), oma=c(0,0,0.4,0))
plot(p, edge_panel = ep, terminal_panel = tp,
     tp_args = list(FUN = identity, width = 15), 
     ip_args = list(fill = paltrees[c(1, 2)]),
     drop_terminal = TRUE, tnex = 1)
@
\endminipage
\minipage{0.45\textwidth}
<<dgp_tree2, fig=TRUE, echo=FALSE, width=4, height=4>>=
par(mar = c(4,4,2,4))
plot(x = d$x[d$z2<=xi], y = d$y[d$z2<=xi], ylim = c(-5,5), ylab = "y", xlab = "x",
     pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
abline(coef = c(0,1), col = pal_dgp[1], lwd = 2)
abline(coef = c(-1,-1), col = pal_dgp[2], lwd = 2)
abline(coef = c(1,-1), col = pal_dgp[3], lwd = 2)
points(x = d$x[d$z2>xi & d$z1<=xi], y = d$y[d$z2>xi & d$z1<=xi], pch = 19, 
       col = alpha(pal_dgp[2], 0.4), cex = 0.7)
points(x = d$x[d$z2>xi & d$z1>xi], y = d$y[d$z2>xi & d$z1>xi], pch = 19, 
       col = alpha(pal_dgp[3], 0.4), cex = 0.7)
legend("top", c(expression(z[2]<=xi), 
                expression(z[2]>xi~" and "~z[1]<=xi),
                expression(z[2]>xi~" and "~z[1]>xi)), 
       col = c(pal[1], pal[2], pal_dgp[3]), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==0), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 2)
mtext(expression(beta[1]==+delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 1.3)
mtext(expression(beta[0]==+delta), side = 4, col = pal_dgp[3], las = 1, line = 0.5, at = 0.1)
mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[3], las = 1, line = 0.5, at = -0.6)
mtext(expression(beta[0]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -2)
mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -2.7)
@
\endminipage
\caption{\label{fig:dgp_tree}Left panel: True tree structure for the second simulation scenario 
with the true model parameters $\beta_0$ and $\beta_1$ for each terminal panel.
Right panel: Bivariate plot of the response $Y$ on regressor $X$ 
with effect size $\delta = 1$.}
\end{center}
\end{figure}

@@

Looking at the results for the split point being close to the upper edge of the considered
interval ($\xi=0.8$, lower panels) all of the applied strategies perform worse than for the
central split point. In this case MOB is clearly ahead leading to the best results. CTree 
performs second best while both GUIDE versions struggle to detect the correct split variable 
even for a high effect size. 
(In order to show that the here observed advantage of MOB over CTree is mainly due to the 
abrupt shift in the model parameters $\beta_0$ and $\beta_1$ an additional evaluation is 
presented in Appendix~\ref{app:contbeta} where these parameters change continuously.)
